Versioned data: why it’s needed and how it can be achieved (easily and cheaply)

### 2017-07-19 ms.tex
work on datastorr section

\mstype{Article Note}
\runninghead{Lightweight versioned data}
\keywords{}

@@

To overcome the limitations of static datasets, many research groups have
setup dedicated webservers to host and distribute data. In these systems,
datasets areas maintained in a structure database, such as SQL, and a front
end to this database is written, often using a platform such as Ruby on rails.
When designed well, such systems are very useful for updating or adding to a
dataset. However, the infrastructure investment (designing maintaining, and
hosting the database) is clearly overkill for most projects and is a very
substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many groups. Moreover, the existence of a live webserver,
does not in itself guarantee ongoing access to different versions of a
dataset.

When considering the right tool to store and deliver versioned data, the size
of the dataset is another essential consideration. ``Big data'' brings with it
specific needs and so requires a specific set of hardware and software tools.
Examples of big data include genetic datasets, global datasets of species
observations, and remote sensing data. This size of this data is bigger than
that which individual research groups can maintain and often requires a
specialized governmental or non-governmental institution with full-time staff
to curate and maintain those research resources. We were particularly
interested in smaller and more specialized problems--the type of datasets that
a given research group might specialize in. In most of these cases, the data
collected will not be ``big'' but rather small to medium sized. Such data
support research projects, but are not in most cases general enough to
establish infra-structure for. But crucially the data themselves will change
over time. And currently there is no infrastructure for organizing and
distributing these types of ``living datasets''.

@@

@@

While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform. Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset. For example, the package
\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples). To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API. With \texttt{datastorr}, setting up a new data-specific package is straightforward and requires minimal computational skills.

>>>>

\mstype{Article Note}
\runninghead{Versioned data delivery}
\keywords{}

@@

To overcome the limitations of static datasets, many research groups have
setup dedicated webservers to host and distribute data. In these systems,
datasets areas maintained in a structure database, such as SQL, and a front
end to this database is written, often using a platform such as Ruby on rails.
When designed well, such systems are very useful for updating or adding to a
dataset. However, the infrastructure investment (designing maintaining, and
hosting the database) is clearly overkill for most projects and is a very
substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
does not in itself guarantee ongoing access to different versions of a
dataset.

When considering the right tool to store and deliver versioned data, the size
of the dataset is another essential consideration. ``Big data'' brings with it
specific needs and so requires a specific set of hardware and software tools.
Examples of big data include genetic datasets, global datasets of species
observations, and remote sensing data. This size of this data is bigger than
that which individual research groups can maintain and often requires a
specialized governmental or non-governmental institution with full-time staff
to curate and maintain those research resources. Here we are particularly
interested in smaller and more specialized problems (e.g., cases where a single 
research group contributes most of the data). In most of these cases, the data
collected will not be ``big'' but rather small to medium sized. Such data
support research projects, but are not in most cases general enough to
establish infra-structure for. But crucially the data themselves will change
over time. And currently there is no infrastructure for organizing and
distributing these types of ``living datasets''.

@@

MP: I think these sections on git would benefit from being turned into paragraphs

@@

GitHub can be used right ``out of the box'' to created versioned data releases
(ideally following semantic versioning standards). Both database contributors and end users
can go onto the website and up- or download the latest (or a previous) version of a dataset.
However, for many use cases, this workflow is not really that efficient. Ideally, researchers
should be able to access the database programmatically. To this end, we offer a novel solution.
We have written a R package, \texttt{datastorr}, that a) interacts with the GitHub API to pull 
down versions of the dataset; and b) constructs the shell of a second, database specific R
package, which can be used to access a given dataset. Using this system, a researcher can create
and distribute a R package that facilitates access to her data with (very) minimal computational skills.

%While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform. Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset. For example, the package
%\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples). To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API. With \texttt{datastorr}, setting up a new data-specific package is straightforward and requires minimal computational skills.
